A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-12-11 papers/water-SAFT/paper.tex
more edits to water paper Tests skipped!

One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature,
stating that
\begin{equation}
  \Omega (T) = \underset{n(\rr)}{min}\left\{ F[n(\rr),T] + \int (V_\textit{ext}(\rr) +\mu)n(\rr)
d\rr\right\},
\end{equation}
where $\Omega (T)$ is the grand potential of a system in the
external potential $V_\textit{ext}$ at temperature $T$, $n(\rr)$ is
the density of atoms or molecules, and $F[n(\rr),T]$ is a universal
free-energy functional for the fluid, which is independent of the
external potential $V_\textit{ext}$.  Classical DFT is a natural
framework for creating a more flexible theory of hydrophobicity that
can readily describe interaction of water with arbitrary external
potentials---such as potentials describing strong interactions with
solutes or surfaces.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination,
  sundararaman2012computationally}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producing a
high value for pressure.
A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.
%
We should give special mention to Sundararaman \emph{et al} recently
introduced a classical density functional for water using an model in
which a water molecule is seen as a hard sphere attached to two
tetrahedrally oriented hard spheres representing voids, or
orientations in which a hydrogen bond may not be formed, with all
attractive interactions being lumped into a single pair potential
treated in a mean field
approximation\cite{sundararaman2012computationally}.

@@

Functionals constructed using a local density approximation fail to
satisfy the contact-value theorem, and therefore incorrectly model
small hard solutes.  The contact-value theorem relates the pressure on
a hard wall interface to the contact density of the fluid at that
interface:
\begin{equation}\label{eq:contact}
  p = n_ck_BT,
\end{equation}
where $n_c$ is the contact density, $k_B$ is the Boltzmann constant
and $T$ is the temperature of the fluid.  The pressure in this formula
is defined as the force per unit \emph{solvent accessible area}, which
differs from the actual pressure when the surface is curved.  For a
cavity in which the solvent is excluded from an arbitrarily small
volume, the contact density will be the same as the bulk density, and
therefore we can integrate the above pressure to find that the excess
chemical potential of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V.
\end{equation}
The contact-value theorem is violated by local classical density
functionals such as those using a local density approximation or a
square-gradient term, but is satisfied by non-local classical density
functionals, such as those using a weighted-density approach.

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
dimensionless parameter $\lscale$---with a fitted value of
0.353---which determines the length scale over which the density is
averaged when computing the dispersion free energy and its derivative.
We determine this final parameter by fitting the computed surface
tension to the experimental surface tension with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of SAFT-based functionals that do not explicitly treat
the critical point---we cannot reasonably describe the surface tension
at all temperatures, and choose to fit the surface tension at and
around room temperature.  We note here that although we could have
chosen to fit the surface tension with a square-gradient term in the
free energy, instead of adjusting the length scale for the dispersive
attraction.  However, would have resulted in a functional that
violates the contact-value theorem, which among other problems would
fail to satisfy Equation~\ref{contactvaluethm} for the excess chemical
potential of small solutes.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent-accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.  For rods with radius larger than 0.5~nm or so, this
reaches a maximum value of 75~mN/m, which is slightly higher than
macroscopic surface tension.  In the limit of very large rods, this
value will decrease and approach the macroscopic value.  As seen in
the inset of Fig.~\ref{fig:energy-vs-diameter}, for rods with very
small radius (less than about 0.5~\AA) the excess chemical potential
is proportional to volume, satisfying Equation~\ref{contactvaluethm},
which results from the contact-value theorem.

@@

Walther \emph{et al}\cite{walther2004hydrophobic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods, using molecular dynamics with the SPC
model for water, for nanotubes of diameter 1.25~nm and separations
ranging from about 0.3~nm to 1.5~nm.  In agreement with our findings
for two purely hydrophobic rods, Walther \emph{et al} find that in the
absence of Lennard-Jones attraction between carbon and oxygen, there
is a drying transition at a distance comparable to the diameter of the
nanotube.  In contrast to this, when the attraction between nanotubes
and water is turned on, they find that the drying transition occurs at
much shorter distances, comparable to the diameter of water.

@@

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the ratio of the excess chemical potential of the cavity
system to the solvent-accessible surface area
(Figure~\ref{fig:sphere-energy-vs-diameter}).  This
effective surface tension surpasses the macroscopic surface tension at a
radius of almost 1~nm, and at large radius will drop to the macroscopic value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
macroscopic surface tension of water at room
temperature\cite{vega2007surface}.

>>>>

One promising approach for an efficient continuum description of water
is that of classical density-functional theory (DFT), which is an
approach for evaluating the free energy and thermally averaged density
of fluids in an arbitrary external potential~\cite{ebner1976density}.
The foundation of classical DFT is the Mermin
theorem\cite{mermin1965thermal}, which extends the Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} to non-zero temperature,
stating that
\begin{equation}
  \Omega (T) = \underset{n(\rr)}{min}\left\{ F[n(\rr),T] + \int (V_\textit{ext}(\rr) +\mu)n(\rr)
d\rr\right\},
\end{equation}
where $\Omega (T)$ is the grand potential of a system in the external
potential $V_\textit{ext}$ at temperature $T$, $n(\rr)$ is the density
of atoms or molecules, $\mu$ is the chemical potential and
$F[n(\rr),T]$ is a universal free-energy functional for the fluid,
which is independent of the external potential $V_\textit{ext}$.
Classical DFT is a natural framework for creating a more flexible
theory of hydrophobicity that can readily describe interaction of
water with arbitrary external potentials---such as potentials
describing strong interactions with solutes or surfaces.

@@

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, yang1994density, gloor2002saft,
  gloor2004accurate, gloor2007prediction, Jaqaman2004,
  clark2006developing, chuev2006, lischner2010classical,
  fu2005vapor-liquid-dft,kiselev2006new, blas2001examination,
  sundararaman2012computationally}, each of
which captures some of the qualitative behavior of water.  However,
each of these functionals also fail to capture some of water's unique
properties.  For instance, the functional of Lischner \emph{et
  al}\cite{lischner2010classical} treats the surface tension
correctly, but can only be used at room temperature, and thus captures
none of the temperature-dependence of water.  A functional by Chuev
and Skolov\cite{chuev2006} uses an ad hoc modification of FMT that can
predict hydrophobic hydration near temperatures of 298~K, but does not
produce a correct equation of state due to their method producing a
high value for pressure.
A number of classical density functionals
have recently been produced that are based on Statistical Associating
Fluid Theory (SAFT)\cite{ segura1997associating, segura1998comparison,
  yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft,gloor2002saft,muller2001molecular,
  clark2006developing, gloor2007prediction, gloor2004accurate,
  gross2009density, kahl2008modified, blas2001examination}.  These
functionals are based on a perturbative thermodynamic expansion, and
do reproduce the temperature-dependence of water's properties.
% 
We should give special mention to Sundararaman \emph{et al} who
recently introduced a classical density functional for water using a
model in which a water molecule is treated as a hard sphere attached to
two tetrahedrally oriented hard spheres representing voids, or
orientations in which a hydrogen bond may not be formed, with all
attractive interactions being lumped into a single pair potential
treated in a mean field
approximation\cite{sundararaman2012computationally}.

@@

Functionals constructed using a local density approximation fail to
satisfy the contact-value theorem, and therefore incorrectly model
small hard solutes.  The contact-value theorem relates the pressure on
a hard surface to the contact density of the fluid at that
surface:
\begin{equation}\label{eq:contact}
  p(\rr_c) = n(\rr_c)k_BT,
\end{equation}
where $\rr_c$ is the position at which a molecule is in contact with
the hard surface, $n(\rr_c)$ is the density at that point of contact,
and $p(\rr_c)$ is the pressure that the fluid exerts on the surface at
the same point.  This pressure is defined as a ratio of force to
\emph{solvent accessible surface} area.
%
For a solute which excludes the solvent from an arbitrarily small
volume, the contact density will be the same as the bulk density, and
therefore we can integrate the above pressure to find that the excess
chemical potential of a small hard solute is proportional to the
solvent-excluded volume:
\begin{equation}\label{contactvaluethm}
  F = n k_BT V.
\end{equation}
The contact-value theorem is violated by local classical density
functionals such as those using a local density approximation or a
square-gradient term, but is satisfied by non-local classical density
functionals, such as those using a weighted-density approach.

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
dimensionless parameter $\lscale$---with a fitted value of
0.353---which determines the length scale over which the density is
averaged when computing the dispersion free energy and its derivative.
We determine this final parameter by fitting the computed surface
tension to the experimental surface tension with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of SAFT-based functionals that do not explicitly treat
the critical point---we cannot reasonably describe the surface tension
at all temperatures, and choose to fit the surface tension at and
around room temperature.  We note here that we could have chosen to
fit the surface tension with a square-gradient term in the free energy
instead of adjusting the length scale for the dispersive attraction.
This would result in a functional that violates the contact-value
theorem which, among other problems, would fail to satisfy
Equation~\ref{contactvaluethm} for the excess chemical potential of
small solutes.

@@

We begin by studying a single hydrophobic rod immersed in water.
% We
% model the density on a grid with spacing of about 0.01~nm, which gives
% the free energy of hydration to a precision of better than a percent,
% and the density at contact to a few percent precision.
In Figure~\ref{fig:energy-vs-diameter} we show the excess chemical
potential at room temperature, scaled by the solvent accessible
surface area of the hard rod, plotted as a function of hard-rod
radius.  We define the hard-rods radius as the radius from which water
is excluded.  For rods with radius larger than 0.5~nm or so, this
reaches a maximum value of 75~mN/m, which is slightly higher than
macroscopic surface tension.  In the limit of very large rods, this
value will decrease and approach the macroscopic value.  As seen in
the inset of Fig.~\ref{fig:energy-vs-diameter}, for rods with very
small radius (less than about 0.5~\AA) the excess chemical potential
is proportional to volume, satisfying Equation~\ref{contactvaluethm},
which results from the contact-value theorem.

@@

Walther \emph{et al}\cite{walther2004hydrophobic} studied the
interactions between two carbon nanotubes, which are geometrically
similar to our hydrophobic rods, using molecular dynamics with the SPC
model for water.  Their simulations used nanotubes of diameter 1.25~nm
and separations ranging from about 0.3~nm to 1.5~nm.  In agreement
with our findings for two purely hydrophobic rods, Walther \emph{et
  al} find that in the absence of Lennard-Jones attraction between
carbon and oxygen, there is a drying transition at a distance
comparable to the diameter of the nanotube.  In contrast to this, when
the attraction between nanotubes and water is turned on, they find
that the drying transition occurs at much shorter distances,
comparable to the diameter of water.

@@

A common model of hydrophobic solutes is the hard-sphere solute, which
is the simplest possible solute, and serves as a test case for
understanding of hydrophobic solutes in
water\cite{sedlmeier2011entropy}.  As in the single rod, we begin by
examining the ratio of the excess chemical potential of the cavity
system to the solvent accessible surface area
(Figure~\ref{fig:sphere-energy-vs-diameter}).  This
effective surface tension surpasses the macroscopic surface tension at a
radius of almost 1~nm, and at large radius will drop to the macroscopic value.  As
with the single rod, we see the analytically correct behavior in the
limit of small solutes.  For comparison, we plot the free energy
calculated using a molecular dynamics simulation of SPC/E
water\cite{huang2001shs}.  The agreement is quite good, apart from the
issue that the SPC/E model for water significantly underestimates the
macroscopic surface tension of water at room
temperature\cite{vega2007surface}.
